---
title: "Supplementary Material for Sluis et al. 2022"
output:
  bookdown::html_document2:
    toc: true
  bookdown::pdf_document2: default
  bookdown::word_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = F)
library(here)
library(tidyverse)
library(broom)
library(ggfortify)
source(here("scripts/data-cleaning.R"))
source(here("scripts/data-wrang.R"))
source(here("scripts/wls-models.R"))
```

R Session info:

```{r r-session}
session <- sessionInfo()
print(session, locale = FALSE)
```

Age group definitions:

| Age category | Age range |
|:---:|:---:|
| 1 | 0 -- 19  |           
| 2 | 20 -- 29 |
| 3 | 30 -- 39 |
| 4 | 40 -- 49 |
| 5 | 50 -- 59 |
| 6 | 60 -- 69 |
| 7 | 70 -- 79 |
| 8 | 80 -- 89 |

# Summary statistics

## Age-at-death within the sample

```{r age-descriptives}
# Age
  # descriptive statistics for whole sample
age_summ_tot <- region_means_all %>%
  filter(region == "all", 
         method == "snodgrass") %>%
  count(id, known_age) %>%
  summarise(n = n(),
            mean = mean(known_age, na.rm = T),
            sd = sd(known_age, na.rm = T),
            range = paste(range(known_age, na.rm = T), collapse = "--")) %>%
  mutate(sex = "Both", .before = mean)
  # descriptive statistics by age category and sex
age_group_summ <- region_means_all %>%
  group_by(age_group, sex) %>%
  count(id, known_age) %>%
  summarise(n = n(),
            mean = mean(known_age, na.rm = T), # mean age within each age cat
            sd = sd(known_age, na.rm = T))

age_sex_summ <- region_means_all %>%
  group_by(sex) %>%
  count(id, known_age) %>%
  summarise(n = n(),
            mean = mean(known_age, na.rm = T),
            sd = sd(known_age, na.rm = T),
            range = paste(range(known_age, na.rm = T), collapse = "--")) %>%
  add_row(age_summ_tot)

females <- age_sex_summ %>%
  filter(sex == "F")
males <- age_sex_summ %>%
  filter(sex == "M")
```

The age distribution between males and females is very similar in mean and shape.
The mean age of males is `r signif(males$mean, 3)` $\pm$ `r signif(males$sd, 3)`, 
with a range of `r males$range`. The mean age of females is `r signif(females$mean, 3)`
$\pm$ `r signif(females$sd, 3)`, with a range of `r females$range`.

```{r age-stats-tbl, echo=FALSE}
knitr::kable(age_sex_summ, caption = "Summary statistics for age-at-death.")
```

```{r expand-age-stats-tbl, echo=FALSE}
knitr::kable(age_group_summ, caption = "Summary statistics for grouped ages-at-death.")
```

```{r age-dist-plot, fig.cap="Age distribution for each sex. F = Female, M = Male."}
raw_data %>%
  filter(vertebral_position == "Superior") %>%
  ggplot(aes(x = known_age, fill = sex)) +
    geom_density(alpha = 0.5) +
    scale_fill_viridis_d() +
  theme_bw()
```

## Osteophyte scores

A comparison of osteophyte scores between methods.

```{r sex-osteophyte-pl, warning=FALSE, fig.cap="Comparing osteophyte scores between sex, separated by region and method."}
region_means_all %>%
  ggplot(aes(x = sex, y = region_mean, fill = sex)) +
    geom_violin(alpha = 0.8) +
    geom_boxplot(width = 0.2, fill = "white") +
    facet_wrap(~ method + region) +
    coord_flip() +
    scale_fill_viridis_d() +
    theme_bw() +
    theme(legend.position = "none")
```


# Correlation tests

```{r cor-sex}
cor_test <- region_means_all %>%
  filter(region == "all", sex == "F", method == "snodgrass") %>%
  dplyr::select(region_mean, known_age)
cor_f <- cor.test(cor_test$region_mean, cor_test$known_age) %>%
  tidy()

cor_test <- region_means_all %>%
  filter(region == "all", sex == "M", method == "snodgrass") %>%
  dplyr::select(region_mean, known_age)
cor_m <- cor.test(cor_test$region_mean, cor_test$known_age) %>%
  tidy()

cor_f %>%
  bind_rows(cor_m) %>%
  mutate(sex = c("F", "M"), .before = 1) %>%
  knitr::kable(caption = "Pearson correlation coefficients for Snodgrass osteophyte score for combined regions, and age-at-death. Separated by Sex.")
```

Males (`r cor_m$estimate`) have a higher correlation between osteophyte score 
and age than females (`r cor_f$estimate`). Both have strong correlations.

# Weighted least squares regression

Models were built using weights estimated with a standard deviation function 
regressing the absolute values of the residuals on the predictor, i.e. known age.
The weights were then calculated as the inverse of the squared standard deviation
function.

$$ w_i = \frac{1}{\sigma^2_i} $$

```{r wls-pl, warning=FALSE, message=FALSE, fig.cap="Regression plots for all regions and methods. F = Female, M = Male."}
# Code to generate wls regression lines for facet plot
reg_plots <- region_means_all %>%
  #group_by(feat, known_age, sex, region, method) %>%
  #summarise(region_mean = mean(score, na.rm = T)) %>%
  ggplot(aes(x = region_mean, known_age)) +
    geom_point(aes(col = sex)) +
    scale_color_viridis_d(name = "Sex") +
    facet_wrap(~ method + region) +
    theme_bw() +
    labs(x = "Mean Osteophyte Score", y = "Age-at-Death")

# Insert coefficients into table
snodgrass_coefs <- lapply(snodgrass_wls, coef) %>%
  unlist() %>%
  matrix(byrow = T, nrow = 4, ncol = 2) 
watanabe_coefs <- lapply(watanabe_wls, coef) %>%
  unlist() %>%
  matrix(byrow = T, nrow = 4, ncol = 2)
praneatpolgrang_coefs <- lapply(praneatpolgrang_wls, coef) %>%
  unlist() %>%
  matrix(byrow = T, nrow = 4, ncol = 2)
all_coefs <- rbind(praneatpolgrang_coefs, snodgrass_coefs, watanabe_coefs)
# Create data frame to pass to geom_abline
reg_coefs <- data.frame("method" = c(rep("praneatpolgrang", 4),
                                     rep("snodgrass", 4), 
                                     rep("watanabe", 4)),
                        "region" = rep(c("all", "cervical", "lumbar", "thoracic")), 
                        "slope" = all_coefs[,1], 
                        "intercept" = all_coefs[,2])
reg_plots +
    geom_abline(aes(slope = slope, intercept = intercept), reg_coefs)
ggsave(here("fig_output", "wls_plots.jpg"),
      dpi = 600, height = 6, width = 7.5, units = "in")
```

## Snodgrass

```{r results='asis'}
lapply(snodgrass_wls, extract_eq, use_coefs = T)
```

## Watanabe et al.

```{r results='asis'}
lapply(watanabe_wls, extract_eq, use_coefs = T)
```

## Praneatpolgrang et al.

```{r results='asis'}
lapply(praneatpolgrang_wls, extract_eq, use_coefs = T)
```


## Model evaluation

Models from each of the three methods (Snodgrass, Watanabe et al., and Praneatpolgrang et al.) 
and four regions (combined, cervical, thoracic, and lumbar) were evaluated using
R$^2$, standard error of the estimate (SEE), AIC, and accuracy.
Accuracy is calculated as the total number of prediction intervals that contain 
the known age, and total number of estimates that are within 5, 10, and 15 years
of the known age.

```{r}
all_eval_tbl %>%
  relocate(method, .before = 1) %>%
  knitr::kable(caption = "Evaluation of combined regions models.")
```

```{r}
knitr::kable(cerv_eval_tbl, caption = "Evaluation of cervical region models.")
```

```{r}
knitr::kable(thor_eval_tbl, caption = "Evaluation of thoracic region models.")
```

```{r}
knitr::kable(lumb_eval_tbl, caption = "Evaluation of lumbar region models.")
```

### Diagnostic plots for the combined regions

**Snodgrass**

```{r diag-pl-snodgrass}
autoplot(snodgrass_wls$all)
```

**Watanabe**

```{r diag-pl-watanabe}
autoplot(watanabe_wls$all)
```

**Praneatpolgrang**

```{r diag-pl-praneatpolgrang}
autoplot(praneatpolgrang_wls$all)
```

# Biases

Does the number of elements present influence accuracy?

Graphs show the number of elements and the absolute value of the residuals for
the Snodgrass method, to see if there is any relationship between the number of 
elements scored and the accuracy of the age estimate.

```{r elements-res-all, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Number of elements (n) plotted against the absolute value of the residuals for Snodgrass method, combined regions."}
n_elements <- element_data_long %>% 
  group_by(id,method,region) %>%
  summarise(n = sum(!is.na(score))) %>%
  ungroup()

n_elements_all <- n_elements %>%
  group_by(id, method) %>%
  summarise(n = sum(n)) %>%
  mutate(region = "all") %>%
  ungroup()

n_all_snodgrass <- n_elements_all %>%
  filter(method == "snodgrass") %>%
  mutate(residual = snodgrass_all_res$residual,
         age = snodgrass_all_res$known_age,
         sex = snodgrass_all_res$sex)

n_all_snodgrass %>%
  ggplot(aes(x = n, y = abs(residual))) +
    geom_point(aes(col = sex)) +
    geom_smooth(method = "lm", se = F, col = "black", size = 0.5) +
    scale_color_viridis_d() +
    theme_minimal()
```

The graph shows that there is no apparent relationship between the number of elements
scored and the accuracy of the age, meaning there is no theoretical minimum
number of elements that needs to be scored; however, I would recommend that at least
3 elements are scored (preferably more) to obtain reliable results.

```{r elements-res-cerv, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Number of elements (n) plotted against the absolute value of the residuals for Snodgrass method, cervical region."}
n_elements <- element_data_long %>% 
  group_by(id,method,region) %>%
  summarise(n = sum(!is.na(score))) %>%
  ungroup()

n_elements_cerv <- n_elements %>%
  filter(region == "cervical") %>%
  group_by(id, method) %>%
  summarise(n = sum(n)) %>%
  #mutate(region = "cervical") %>%
  ungroup()

n_cerv_snodgrass <- n_elements_cerv %>%
  filter(method == "snodgrass", n != 0) %>%
  mutate(residual = snodgrass_cerv_res$residual,
         age = snodgrass_cerv_res$known_age,
         sex = snodgrass_cerv_res$sex)

n_cerv_snodgrass %>%
  ggplot(aes(x = n, y = abs(residual))) +
    geom_point(aes(col = sex)) +
    geom_smooth(method = "lm", se = F, col = "black", size = 0.5) +
    scale_color_viridis_d() +
    theme_minimal()
```

```{r elements-res-thor, eval=TRUE, message=FALSE, warning=FALSE, , fig.cap="Number of elements (n) plotted against the absolute value of the residuals for Snodgrass method, thoracic region."}
n_elements <- element_data_long %>% 
  group_by(id,method,region) %>%
  summarise(n = sum(!is.na(score))) %>%
  ungroup()

n_elements_thor <- n_elements %>%
  filter(region == "thoracic") %>%
  group_by(id, method) %>%
  summarise(n = sum(n)) %>%
  #mutate(region = "thoracic") %>%
  ungroup()

n_thor_snodgrass <- n_elements_thor %>%
  filter(method == "snodgrass", n != 0) %>%
  mutate(residual = snodgrass_thor_res$residual,
         age = snodgrass_thor_res$known_age,
         sex = snodgrass_thor_res$sex)

n_thor_snodgrass %>%
  ggplot(aes(x = n, y = abs(residual))) +
    geom_point(aes(col = sex)) +
    geom_smooth(method = "lm", se = F, col = "black", size = 0.5) +
    scale_color_viridis_d() +
    theme_minimal()
```

```{r elements-res-lumb, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Number of elements (n) plotted against the absolute value of the residuals for Snodgrass method, lumbar region."}
n_elements <- element_data_long %>% 
  group_by(id,method,region) %>%
  summarise(n = sum(!is.na(score))) %>%
  ungroup()

n_elements_lumb <- n_elements %>%
  filter(region == "lumbar") %>%
  group_by(id, method) %>%
  summarise(n = sum(n)) %>%
  #mutate(region = "lumbar") %>%
  ungroup()

n_lumb_snodgrass <- n_elements_lumb %>%
  filter(method == "snodgrass", n != 0) %>%
  mutate(residual = snodgrass_lumb_res$residual,
         age = snodgrass_lumb_res$known_age,
         sex = snodgrass_lumb_res$sex)

n_lumb_snodgrass %>%
  ggplot(aes(x = n, y = abs(residual))) +
    geom_point(aes(col = sex)) +
    geom_smooth(method = "lm", se = F, col = "black", size = 0.5) +
    scale_color_viridis_d() +
    theme_minimal()
```

## Age bias

```{r snodgrass-all-res, fig.cap="Residuals for Snodgrass method, combined regions, plotted against known age-at-death."}
data.frame(
  "known_age" = sep_regions$snodgrass$all$known_age,
  "residuals" = snodgrass_wls$all$residuals,
  "sex" = sep_regions$snodgrass$all$sex) %>%
ggplot(aes(x = known_age, y = residuals, col = sex)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(y = "Residuals", x = "Age-at-Death") +
  scale_colour_viridis_d(name = "Sex")
```

```{r include=FALSE}
ggsave(here("fig_output", "snod_age-bias.jpg"), 
      dpi = 600, width = 7, height = 6, units = "in")
```

```{r snodgrass-cerv-res, fig.cap="Residuals for Snodgrass method, cervical region, plotted against known age-at-death."}
data.frame(
  "known_age" = na.omit(sep_regions$snodgrass$cervical)$known_age,
  "residuals" = snodgrass_wls$cervical$residuals) %>%
ggplot(aes(x = known_age, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_minimal()
```

```{r snodgrass-thor-res, fig.cap="Residuals for Snodgrass method, thoracic region, plotted against known age-at-death."}
data.frame(
  "known_age" = na.omit(sep_regions$snodgrass$thoracic)$known_age,
  "residuals" = snodgrass_wls$thoracic$residuals) %>%
ggplot(aes(x = known_age, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_minimal()
```

```{r snodgrass-lumb-res, fig.cap="Residuals for Snodgrass method, lumbar region, plotted against known age-at-death."}
data.frame(
  "known_age" = na.omit(sep_regions$snodgrass$lumbar)$known_age,
  "residuals" = snodgrass_wls$lumbar$residuals) %>%
ggplot(aes(x = known_age, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_minimal()
```

```{r watanabe-all-res, , fig.cap="Residuals for Watanabe method, combined regions, plotted against known age-at-death."}
data.frame(
  "known_age" = sep_regions$watanabe$all$known_age,
  "residuals" = watanabe_wls$all$residuals,
  "sex" = sep_regions$watanabe$all$sex) %>%
ggplot(aes(x = known_age, y = residuals, col = sex)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(y = "Residuals", x = "Age-at-Death") +
  scale_colour_viridis_d(name = "Sex")
```

```{r include=FALSE}
ggsave(here("fig_output", "watanabe_age-bias.jpg"), 
      dpi = 600, width = 7, height = 6, units = "in")
```

```{r praneat-all-res, fig.cap="Residuals for Praneatpolgrang method, combined regions, plotted against known age-at-death."}
data.frame(
  "known_age" = sep_regions$praneatpolgrang$all$known_age,
  "residuals" = praneatpolgrang_wls$all$residuals,
  "sex" = sep_regions$praneatpolgrang$all$sex) %>%
ggplot(aes(x = known_age, y = residuals, col = sex)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw() +
  labs(y = "Residuals", x = "Age-at-Death") +
  scale_colour_viridis_d(name = "Sex")
```

```{r include=FALSE}
ggsave(here("fig_output", "praneat_age-bias.jpg"), 
       dpi = 600, width = 7, height = 6, units = "in")
```

What is the accuracy and average residuals (not squared or absolute) per age
category?

```{r}
snodgrass_all_bias <- snodgrass_all_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            bias_snodgrass = mean(residual))

watanabe_all_bias <- watanabe_all_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            bias_watanabe = mean(residual))

praneatpolgrang_all_bias <- praneatpolgrang_all_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            bias_praneat = mean(residual))

all_bias <- list(snodgrass_all_bias, watanabe_all_bias, praneatpolgrang_all_bias)

all_mean_bias <- all_bias %>%
  reduce(inner_join, by = c("age_group", "n"))

knitr::kable(all_mean_bias, caption = "Prediction bias by age for combined regions")
```


```{r}
snodgrass_cerv_bias <- snodgrass_cerv_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            snodgrass_bias = mean(residual))

watanabe_cerv_bias <- watanabe_cerv_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            watanabe_bias = mean(residual))

praneatpolgrang_cerv_bias <- praneatpolgrang_cerv_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            praneat_bias = mean(residual))

cerv_bias <- list(snodgrass_cerv_bias, watanabe_cerv_bias, praneatpolgrang_cerv_bias)

cerv_mean_bias <- cerv_bias %>%
  reduce(inner_join, by = c("age_group", "n"))

knitr::kable(cerv_mean_bias, caption = "Prediction bias by age for cervical region")
```

```{r}
snodgrass_thor_bias <- snodgrass_thor_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            snodgrass_bias = mean(residual))

watanabe_thor_bias <- watanabe_thor_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            watanabe_bias = mean(residual))

praneatpolgrang_thor_bias <- praneatpolgrang_thor_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            praneat_bias = mean(residual))

thor_bias <- list(snodgrass_thor_bias, watanabe_thor_bias, praneatpolgrang_thor_bias)

thor_mean_bias <- thor_bias %>%
  reduce(inner_join, by = c("age_group", "n"))

knitr::kable(thor_mean_bias, caption = "Prediction bias by age for thoracic region")
```

```{r}
snodgrass_lumb_bias <- snodgrass_lumb_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            snodgrass_bias = mean(residual))

watanabe_lumb_bias <- watanabe_lumb_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            watanabe_bias = mean(residual))

praneatpolgrang_lumb_bias <- praneatpolgrang_lumb_res %>%
  group_by(age_group) %>%
  summarise(n = n(),
            praneat_bias = mean(residual))

lumb_bias <- list(snodgrass_lumb_bias, watanabe_lumb_bias, praneatpolgrang_lumb_bias)

lumb_mean_bias <- lumb_bias %>%
  reduce(inner_join, by = c("age_group", "n"))

knitr::kable(lumb_mean_bias, caption = "Prediction bias by age for lumbar region")
```



